Dependent Types for Safe and Secure Web Programming

### 2014-01-30 paper/conclusion.tex
Paper revisions

\citet{thiemann:wash} adds the notion of a CGI Session for maintaining
state, and provides more sophisticated form-handling methods. In particular,
callbacks may be associated with submit buttons, with nameless representations
for form inputs. Due to the unavailability of full dependent types in Haskell,
however, this implementation does not statically verify the suitability of the
callback function for the form inputs. Both implementations of the CGI library,
being built upon monads, mean that the use of additional effects such as
database access is achieved either through monad transformers or 
performing arbitrary IO operations. Both of these approaches are limited---the
former does not scale well to multiple effects, and the latter allows for the
introduction of errors by allowing the violation of resource usage protocols.

\citet{plasmeijer:idata} describe an alternative approach to type-safe form
handling through the
\textit{i}nteractive \textit{Data}, or \idata{} abstraction. Instead of
processing being triggered by form submission, as in the approach described in
this paper, applications created in the \idata{} toolkit are 
edit-driven. This means that upon a component being edited, a computation
occurs, given the state of the current form. This is saved for
future computations. Should a user enter invalid data, for example by entering
text in a field designated for integers, the change will be reverted. This is
demonstrated practically through the use of \idata{} to implement a conference
management system \cite{plasmeijer:cms}.

The concept of \idata{} is taken further by the introduction of \itasks{}
\cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple
\idata{} forms to interact with one another. This is achieved using
high-level combinators which allow the implementation of concepts such as
recursion, sequence and choice in a scalable fashion. These high-level
abstractions are elegant, but the style and syntax differ substantially from a
traditional web application. Our approach takes the concept of type-safe input
handling and uses it in a more traditional fashion, whilst retaining the
type-retention guarantees afforded by \idata{} elements.

@@

%Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
%dependent types to enable the generation of provably-correct and secure queries
%and DOM code, without requiring overly complex types or proofs. Through the use
%of the Ur\/Web DSL, developers may generate well-typed code from records. By
%statically verifying the \textit{code generator}, it is therefore possible to
%assert that the generated code is secure. Such ideas are extremely interesting
%for the implementation of object-relational mapping, which we hope to implement
%in the near future.

WebDSL \cite{webdsl} is a domain-specific language written primarily to
introduce new abstractions which aim to reduce the amount of boilerplate code
that must be written and maintained by developeres. WebDSL is built on top of
Java, which often includes a large amount of redundant code such as accessor
and mutator functions within entity classes. The Java Persistence API (JPA)
\cite{jpa} provides an object-relational mapping through the use of Java 5
annotations, which may then be used to construct database tables. These
annotations soon become complex, however, and coupled with redundant
boilerplate code, data model declarations may soon become unwieldy. Through the
use of WebDSL, these data model declarations can be much more compactly
declared, and elaborated into Java code by parsing the data-modelling DSL into
an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL
also applies similar concepts to implement a \textit{template system} for such
objects, which allows the data to be used in code generation. We look to
implement many of these ideas, but as effects within the IdrisWeb
framework, as with the form construction effect.

@@

There are many other applications which make use of specific resource usage
protocols, for example popular libraries such as
\textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
Applying a similar approach would allow for sensitive programs requiring
cryptographic routines to be written using a language with full dependent
types, in turn adding an extra layer of confidence in their security.

@@

Type providers, as originally implemented in F\# \cite{msr:tp}, are an
interesting method by which external data sources may be used to import
external information, such that it may be used during compilation. 
In this way, it becomes possible to use the extra type
information to statically ensure the validity of artefacts such as SQL
queries and data structures. If data structures within the program do not
conform to a given database schema, for example, then the program will not
type-check.  This has been implemented for 
\idris{} 
\cite{christiansen:dtp}, exploiting the fact that types can be
calculated by functions to avoid generating extra code in the type provider
step. Dependent type providers additionally have stronger safety guarantees as
they may not generate unchecked code, but at the same time this is matched by a
decrease in expressiveness. Nonetheless, such techniques provide a promising
mechanism to verify the semantic soundness of programs and we look to
investigate their integration in further work.

>>>>

\citet{thiemann:wash} adds the notion of a CGI Session for maintaining
state, and provides more sophisticated form-handling methods. In particular,
callbacks may be associated with submit buttons, with nameless representations
for form inputs. Both implementations of the CGI library,
being built upon monads, mean that the use of additional effects such as
database access is achieved either through monad transformers or 
performing arbitrary IO operations. Both of these approaches are limited---the
former does not scale well to multiple effects, and the latter allows for the
introduction of errors by allowing the violation of resource usage protocols.

\citet{plasmeijer:idata} describe an alternative approach to type-safe form
handling through the
\textit{i}nteractive \textit{Data}, or \idata{} abstraction. Instead of
processing being triggered by form submission, as in the approach described in
this paper, applications created in the \idata{} toolkit are 
edit-driven. This means that upon a component being edited, a computation
occurs, given the state of the current form. Should a user enter invalid data, for example by entering
text in a field designated for integers, the change will be reverted. This is
demonstrated practically through the use of \idata{} to implement a conference
management system \cite{plasmeijer:cms}.

The concept of \idata{} is taken further by the introduction of \itasks{}
\cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple
\idata{} forms to interact with one another. This is achieved using
high-level combinators which allow the implementation of concepts such as
recursion, sequence and choice in a scalable fashion.

@@

Formlets \cite{cooper:formlets} are a well-known functional abstraction over web forms, making use of McBride and Paterson's applicative functors \cite{mcbride:applicative} to provide an extensible and powerful method of building and handling web forms. 
Our approach differs in that we check that handler functions conform to the types of the elements in the forms by parameterising the resource associated with the form construction effect, as opposed to using a preprocessor to rewrite the form elements in applicative notation.
Our framework does not yet support composition of sections of forms, as is the case with formlets, but we foresee no problems with extending our DSL to add this functionality.

Java is often used as a language to write enterprise-level web applications. Frameworks such as the Java Persistence API (JPA) \cite{jpa} use Java annotations to translate data models into the appropriate database schemata, but such code may soon become unwieldy due to the large amount of redundant boilerplate code such as accessor and mutator functions. WebDSL \cite{webdsl} is a domain-specific language written primarily to
introduce new abstractions which aim to reduce the amount of boilerplate code
that must be written and maintained by developers. 
The DSL is parsed into an abstract syntax tree, modified using rewrite rules, and elaborated back into Java code. WebDSL also applies similar concepts to implement a \textit{template system} for the presentation of data entities. We look to
implement many of these ideas, but as effects within the IdrisWeb
framework, as with the form construction effect.

@@

There are many other applications which make use of specific resource usage
protocols, for example popular cryptography libraries such as
\textit{Sodium}\footnote{\texttt{https://github.com/jedisct1/libsodium}}.
Applying a similar approach would allow for sensitive programs requiring
cryptographic routines to be written using a language with full dependent
types, in turn adding an extra layer of confidence in their security.

@@

Type providers, as originally implemented in F\# \cite{msr:tp}, are an
interesting method by which external data sources may be used to import
external information, such that it may be used during compilation. 
In this way, it becomes possible to use the extra type
information to statically ensure the validity of artefacts such as SQL
queries and data structures. If data structures within the program do not
conform to a given database schema, for example, then the program will not
type-check.  This has been implemented for 
\idris{} 
\cite{christiansen:dtp}, exploiting the fact that types can be
calculated by functions to avoid unsafely generating extra code in the type provider
step. We believe that this technique would be an interesting avenue of exploration to provide further static guarantees about the form of database queries.

### 2014-01-30 paper/protocols.tex
Paper revisions

%A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell. Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program. If this were to happen, then the program would fail due to the fact that headers had not been written to the client.

By modelling CGI as a resource-dependent algebraic effect, we may enforce a
resource usage protocol which
%even though the program may be running in an IO execution context, 
prevents arbitrary IO from being performed and therefore
ensures that the headers are written correctly. We
define an effect, \texttt{Cgi}, and an associated resource,
\texttt{InitialisedCGI}, parameterised over the current state,
\texttt{CGIStep}, and containing a
\texttt{CGIInfo} record which contains information from the request. We
represent an uninitialised CGI process as the unit type, \texttt{()}.

@@

>>>>

By modelling CGI as a resource-dependent algebraic effect, we may enforce a
resource usage protocol which
prevents arbitrary IO from being performed and therefore
ensures that the headers are written correctly. We
define an effect, \texttt{Cgi}, and an associated resource,
\texttt{InitialisedCGI}, parameterised over the current state,
\texttt{CGIStep}, and containing a
\texttt{CGIInfo} record which contains information from the request. We
represent an uninitialised CGI process as the unit type, \texttt{()}.

@@

Here, \texttt{output} is a function which appends some output to the CGI output buffer, which is later written to the page.
